Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2019-08-19 paper/abstract.tex
redrafting

Due to a lack of precise ages and rotation periods for old main-sequence
stars, the rotational evolution of FGKM dwarfs is mostly unconstrained
after around 2-3 billion years.
However, recent evidence suggests that the relationship between rotation
period and color/temperature/mass does not remain constant over large
timescales, contradicting a long-held assumption that underlies many
gyrochronology relations.
In this work we use the kinematic ages of populations of K dwarfs to reveal
the evolving relationship between rotation period and effective temperature.
We find that groups of stars, selected to be the same age, do {\it not} have
the same velocity dispersion at all temperatures; cooler stars have greater
velocity dispersions than hotter stars.
Two different scenarios could be responsible for producing this observation:
either the rotation period-color relation changes over time, or stars of
different masses experience different rates of dynamical heating.
% We show however, that if mass-dependent heating is responsible, the
% differential heating rate is not constant over logarithmic time intervals
% which implies that different heating processes must be acting before and after
% ages of around 1 Gyr.
% Perhaps a more likely scenario is that the relation between rotation period
% and color changes over time, undergoing a particularly sharp transition at
% around 1 Gyr.
If we assume that mass-dependent heating does not affect the data, our results
indicate that the slope of the period-color relation actually {\it inverts}:
at young ages early K stars rotate more rapidly than late K stars, but beyond
$\sim$ 2 Gyr they rotate more slowly.

>>>>

The rotational evolution of cool dwarfs is mostly unconstrained after around
2-3 billion years, and the gyrochronology relations are poorly calibrated
beyond this time due to a lack of precise ages and rotation periods for old
main-sequence stars.
% However, recent evidence suggests that the relationship between rotation
% period and color/temperature/mass does not remain constant over large
% timescales, contradicting a long-held assumption that underlies many
% gyrochronology relations.
In this work we use the velocities of populations of K dwarfs to reveal the
rotational evolution of these cool stars at old ages.
We find that groups of stars, selected to be the same age using a
gyrochronology relation calibrated to the Praesepe cluster, do {\it not} have
the same velocity dispersion at all temperatures; cooler stars have greater
velocity dispersions than hotter stars.
We interpret this to mean that these cooler stars are older than the hotter
ones, which implies that the period-color relation of Praesepe is not
appropriate for old stars.
Futhermore, our results indicate that the slope of the period-color relation
actually {\it inverts}: at young ages early K stars rotate more rapidly than
late K stars, but beyond $\sim$ 2 Gyr they rotate more slowly.
These observations could be reproduced by mass-dependent dynamical heating,
where lower-mass stars experience a greater rate of dynamical heating than
higher-mass stars, however we find no evidence for mass-dependent heating
within K and M dwarfs in the \kepler\ field.

% We show however, that if mass-dependent heating is responsible, the
% differential heating rate is not constant over logarithmic time intervals
% which implies that different heating processes must be acting before and after
% ages of around 1 Gyr.
% Perhaps a more likely scenario is that the relation between rotation period
% and color changes over time, undergoing a particularly sharp transition at
% around 1 Gyr.
% If we assume that mass-dependent heating does not affect the data,

### 2019-08-19 paper/conclusion.tex
redrafting

We found that old groups of K dwarfs selected to be coeval using the
\citet{angus2019} gyrochronology relation do {\it not} have the same velocity
dispersion across all temperatures.
This indicates that either the \citep{angus2019} period-color relation is
wrong, or that mass-dependent dynamical heating is affecting this data set.
We presented three pieces of evidence that suggest mass-dependent heating
cannot be responsible for all of the increasing velocity dispersion.

\begin{itemize}

\item{
        No strong evidence for mass-dependent heating has been found among M, L
and T-type stars.
These low-mass stars live longer than the current age of the Universe and
        therefore do not suffer from a mass-age degeneracy
        \citep{faherty2009}.
}

\item{
        Literature measurements of the (\vz) AVR exponent are slightly higher
than the value measured from our data set, but calculated using more massive
stars.
Mass-dependent dynamical heating should however produce a {\it smaller} AVR
        exponent for higher-mass stars.
We acknowledge that the stars in the \mct\ sample, analyzed in this work, are
        subject to different selection biases than those used to calculate
        AVRs in the literature \citep{holmberg2009, aumer2009, yu2018}, so
        drawing direct comparisons between them does not provide a reliable
        test.
}

\item{
        The differences in the (\vb) AVR exponent between the hottest and
coolest K dwarfs in our sample are extremely large, spanning values between
0.4 and 0.75, yet the difference in mass is less than a factor of two.
Such a large spread in AVR exponents cannot be explained by mass-dependent
        heating alone.
}

\end{itemize}

@@

>>>>

We found that old groups of K dwarfs selected to be coeval using the
\citet{angus2019} gyrochronology relation do {\it not} have the same velocity
dispersion across all temperatures.
% This indicates that either the \citep{angus2019} period-color relation is
% incorrect, or that mass-dependent dynamical heating is affecting this data set.
% We presented three pieces of evidence that suggest mass-dependent heating
% cannot be responsible for all of the increasing velocity dispersion.

% \begin{itemize}

% \item{
%         No strong evidence for mass-dependent heating has been found among M, L
% and T-type stars.
% These low-mass stars live longer than the current age of the Universe and
%         therefore do not suffer from a mass-age degeneracy
%         \citep{faherty2009}.
% }

% \item{
%         Literature measurements of the (\vz) AVR exponent are slightly higher
% than the value measured from our data set, but calculated using more massive
% stars.
% Mass-dependent dynamical heating should however produce a {\it smaller} AVR
%         exponent for higher-mass stars.
% We acknowledge that the stars in the \mct\ sample, analyzed in this work, are
%         subject to different selection biases than those used to calculate
%         AVRs in the literature \citep{holmberg2009, aumer2009, yu2018}, so
%         drawing direct comparisons between them does not provide a reliable
%         test.
% }

% \item{
%         The differences in the (\vb) AVR exponent between the hottest and
% coolest K dwarfs in our sample are extremely large, spanning values between
% 0.4 and 0.75, yet the difference in mass is less than a factor of two.
% Such a large spread in AVR exponents cannot be explained by mass-dependent
%         heating alone.
% }

% \end{itemize}

@@

\racomment{How does the weakened van Saders braking law fit into all this?}

\racomment{Talk about the fact that the stars in each bin are not necessarily
the same age.}

\racomment{Dispersion could be dependent on the selection function.
We cut stars at high b, because we find that these do lead to a bias.
We find no evidence for increased velocity dispersion with decreasing teff in
either vb or vz.
We find that sigma clipping is necessary for vb but not vz and do not find
that the outliers are distributed towards higher galactic latitudes.
}

% \racomment{Trevor suggest that stars forming in clusters could be the reason
% why young stars are so dynamically cool.}

% \racomment{Are there any asteroseismic stars in this sample that could help
% pin down the age?}

### 2019-08-19 paper/intro.tex
less cringy
redrafting

\subsection{Gyrochronology}
It has long been known that the rotation periods of FGKM dwarfs increase over
time \citep{skumanich1972}.
This characteristic of main-sequence stars allows them to be dated via their
rotation periods in a practice known as gyrochronology, which is useful
since the ages of main-sequence stars are extremely difficult to measure via
the traditional age-dating method of isochrone placement.
It is also well established that stars of the same age but different {\it
masses} have different rotation periods \racomment{(citation)}, thought to be
caused by the deeper convective zones, and therefore stronger magnetic dynamos
(and more efficient magnetic braking) in lower-mass stars.
However, an underlying assumption behind many empirical gyrochronology
relations \citep[\eg][]{barnes2003, barnes2007, mamajek2008, meibom2011,
angus2015, angus2019} is that the relationships between rotation period and
photometric color\footnote{As the directly observable quantity, color is
usually used as a mass-proxy, and empirical gyrochronology relations are
usually calibrated in color, rather than mass or effective temperature.
}, and rotation period and age are {\it separable}.
In other words, the period-color relation is the same at all ages and the
period-age relation is the same at all colors.
It was recently shown that old field stars rotate more rapidly than a simple,
separable gyrochronology relation would predict \citep{angus2015,
vansaders2016, vansaders2018, metcalfe2019}, and that a mass-dependent
modification to the classical \citet{skumanich1972} spin-down law is required
to reproduce the data \citep{vansaders2016, vansaders2018}.
An even more recent analysis of middle-aged open clusters provides further
indication that the exponent of the period-age relation is mass-dependent.
New rotation period measurements of the 1.1 Gyr open cluster, NGC 6811,
reveal a flattened relationship between rotation period and color
\citep{curtis2019}.
In this cluster, the G dwarfs rotate at the same rate as the K dwarfs.

Unfortunately, the open clusters with rotation period measurements are mostly
young -- currently the oldest is 2.5 Gyr (NGC 6819).
Without rotation periods for precisely dated old stars, it is extremely
difficult to calibrate the relationship between rotation period and color at
old ages.
For this reason, we used a population-based stellar age indicator, velocity
dispersion, to investigate the period-color relations at old ages.
% Since a separable period-age and period-color relation provided an adequate
% fit for the young stars, it was naturally assumed to apply to old stars too.

@@

Stars are thought to be born in the thin disk of the Milky Way, orbiting the
center of the galaxy with a low out-of-plane, or vertical, velocity ($W$, or
$v_z$).
Young stars have relatively small vertical velocities, but gain momentum in
the vertical direction over time.
Although the cause of orbital heating is not well understood, interactions
with giant molecular clouds and spiral arms are thought to play an important
role \racomment{(citations)}.
Although the velocity of any individual star will only provide a weak age
constraint, the velocity dispersion of a group of stars can indicate whether,
on average, that group is old or young relative to other groups.
In this work we compare the velocity dispersions of groups of stars to
ascertain which groups are older and which younger and draw conclusions based
on the implied relative ages of populations.
% Stellar velocities have a history of being used as an age proxy, with several
% notable examples within stellar astronomy \citep[\eg][]{faherty2009,
% west2011}.
Since the age-velocity dispersion relations (AVRs) are themselves still
actively being calibrated, it is difficult to directly compare gyrochronal
ages with kinematic ones.
However, regardless of the exact relation between velocity dispersion and
stellar age, it is expected to be a monotonic relationship, therefore velocity
dispersion can be used effectively to {\it rank} groups of stars by age.

% Vertical {\it actions} are better age indicators than velocities, because
% actions are calculated by integrating angular momentum over the Milky Way's
% potential, and are therefore position invariant -- \eg\ a star will have the
% same action at periapsis and apoapsis.
% In contrast, orbital velocities are different at periapsis and apoapsis -- so
% in this sense, {\it actions} are the natural quantities to use as age proxies.
Vertical action is a better age indicator than vertical velocity, however both
vertical action ($J_z$) and vertical velocity (\vz/W) can only be calculated
with full 6-dimensional position and velocity information.
Unfortunately, most stars with measured rotation periods do not have radial
velocity (RV) measurements because they are relatively faint \kepler\ targets
($\sim$11th-18th magnitudes).
For this reason, we used an alternative age proxy: velocity in the direction
of galactic latitude, \vb.
The \kepler\ field is positioned at low galactic latitude (b=5-20\degrees), so
\vb\ is a close approximation to \vz.

@@

In this paper, we tested the \citet{angus2019} gyrochronology relation, which
was calibrated using the period-color relation of Praesepe (in \gaia\ \gcolor\
color) and the period-age relation of Praesepe and the Sun.
The large number of Praesepe members with precise rotation periods from the
\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
through early M, makes it the ideal cluster for calibrating the period-color
relation of stars at 650 Myrs.
% This relation was calibrated by fitting a 5th-order polynomial to the relation
% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
% and the Sun.
Although only calibrated using Praesepe ($\sim$ 650 Myr) and the Sun (4.56
Gyr), the \citet{angus2019} gyrochronology relation predicts accurate ages for
members of NGC 6819, a 2.5 Gyr open cluster.
However there are no rotation periods for K or M dwarfs in this cluster -- its
coolest members with rotation periods are G dwarfs.
The period-color relation of Praesepe is nearly identical to the period-color
relation of the Hyades, a cluster of around the same age: $\sim$ 650 years
\citep{douglas2016, douglas2017, rebull2017}.
However, Praesepe and the Hyades do {\it not} have the same period-color
relation as NGC 6811, a 1.1 Gyr cluster \citep{curtis2019}.
The G dwarfs in NGC 6811 rotate at the same rate as the K dwarfs and it
appears as though the K stars have `stalled' -- their spin-down has been
halted.
The \citet{angus2019} gyrochronology relation assumes that the rotation
period-color relation of Praesepe is applicable to stars of all ages: the same
polynomial relation fit to Praesepe is used to describe the period-color
relation for all stars.
However, the NGC 6811 cluster suggests that this assumption does not hold for
K dwarfs past around 1 Gyr.
If the period-color relation of the \citet{angus2019} gyrochronology model
{\it were} a perfect model for the rotational evolution of stars, then groups
of stars selected to be similar ages using this relation should fall on an
isochrone on a CMD.
% and have the same velocity dispersion across all colors.
Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
variations in metallicity and extinction, blur out the main sequence enough
that differences between stars in different age and rotation bins are not
easily discernable on the CMD, although there is still a general age and
rotation period gradient on the CMD, as seen in figures \ref{fig:CMD_cuts} and
\ref{fig:age_gradient}.
For this reason, we chose to use {\it kinematic} `isochrones', instead of
magnitude and color-based isochrones: although kinematics as an age indicator
is not necessarily as well calibrated to an absolute age scale as CMD
position, it can be very sensitive to {\it differences} in the ages of stellar
populations.

Isochrones and stellar evolution tracks are highly dependent on choices made
about input physics and assumptions.
% and have often been calibrated using
% different types of stars.
As a result, different sets of models can have very different shapes on the
CMD, particularly at low masses.
In fact, only empirical models, not physical ones, are currently able to
reproduce the CMD positions of M dwarfs.
Instead of relying on CMD position to age-date groups of stars, we opted to
explore age trends via kinematics.
Kinematic age-dating has the advantage of being relatively model independent,
or at least, having a very simple model: that velocity dispersion increases
over time.
This means that it is relatively easy to rank groups of stars by age: older
groups have a larger velocity dispersion.
% It is less easy to rank-order groups of stars by age on the CMD because
% stellar evolution is highly mass and metallicity dependent, and because
% different stellar evolution models predict different ages for the same CMD
% position.
% This does not mean that it is not possible to calibrate gyrochronology using
% isochrones, however it is not the focus of this study.

>>>>

\subsection{Gyrochronology}
It is well established that the rotation periods of FGKM dwarfs increase over
time \citep{skumanich1972}.
This characteristic of main-sequence (MS) stars allows them to be dated via
their rotation periods in a practice known as gyrochronology, which is useful
since the ages of MS stars are extremely difficult to measure via the
traditional age-dating method of isochrone placement.
It is also well established that stars of the same age but different masses
have different rotation periods \racomment{(citation)}, thought to be caused
by the deeper convective zones, and therefore stronger magnetic dynamos (and
more efficient magnetic braking) in lower-mass stars.
However, an underlying assumption behind many empirical gyrochronology
relations is that the relationships between rotation period and photometric
color\footnote{As the directly observable quantity, color is usually used as a
mass-proxy, and empirical gyrochronology relations are usually calibrated in
color, rather than mass or effective temperature.
}, and rotation period and age are {\it separable}, meaning that the
period-color relation is the same at all ages and the period-age relation is
the same at all colors \citep[\eg][]{barnes2003, barnes2007, mamajek2008,
meibom2011, angus2015, angus2019}.
It was recently shown that old field stars rotate more rapidly than a simple,
separable gyrochronology relation would predict \citep{angus2015,
vansaders2016, vansaders2018, metcalfe2019}, and that a mass-dependent
modification to the classical \citet{skumanich1972} spin-down law is required
to reproduce the data \citep{vansaders2016, vansaders2018}.
An even more recent analysis of the 1.1 Gyr open cluster, NGC 6811, provides
further indication that the exponent of the period-age relation is
mass and time-dependent.
This cluster has a flattened relationship between rotation period and color:
the G dwarfs rotate at the same rate as the K dwarfs \citep{curtis2019}.
% Since a separable period-age and period-color relation provided an adequate
% fit for the young stars, it was naturally assumed to apply to old stars too.

In this paper, we tested the \citet{angus2019} gyrochronology relation, which
was calibrated using the period-color relation of Praesepe (in \gaia\ \gcolor\
color) and the period-age relation of Praesepe and the Sun.
The large number of Praesepe members with precise rotation periods from the
\ktwo\ mission \citep{douglas2017, rebull2017}, spanning spectral types F
through early M, makes it the ideal cluster for calibrating the period-color
relation of stars at 650 Myrs.
% This relation was calibrated by fitting a 5th-order polynomial to the relation
% between (log) rotation period and (log) \gaia\ \gcolor\ color for around 800
% members of the Praesepe cluster, and a straight line in (log) age to Praesepe
% and the Sun.
This relation accurately describes the rotation periods of F and G stars up to
around 2.5 Gyr (the age of NGC 6819 -- the oldest cluster with available
rotation periods), but over-predicts the rotation periods of K dwarfs in the
1.1 Gyr NGC 6811 cluster.
No reliable age estimates exist for K dwarfs with rotation periods older than
1.1 Gyr, so the rotational evolution of these stars beyond this age is a
mystery.
% In this paper, we shed light on the rotational evolution of old K dwarfs using
% population-based dynamical ages.
% Unfortunately, open clusters with rotation period measurements are mostly
% young -- currently the oldest is 2.5 Gyr (NGC 6819).
% Without rotation periods for precisely dated old stars, it is extremely
% difficult to calibrate the relationship between rotation period and color at
% old ages.
For this reason, we used a population-based stellar age indicator, velocity
dispersion, to investigate the period-color relations at old ages.

@@

Stars are thought to be born in the thin disk of the Milky Way (MW), orbiting
the center of the galaxy with a low out-of-plane, or vertical, velocity ($W$,
or $v_z$).
% because the MW's gas disk is thin and its gas moves in circular
% orbits.
Young stars have relatively small vertical velocities, but gain momentum in
the vertical direction over time.
Although the cause of orbital heating is not well understood, interactions
with giant molecular clouds, spiral arms and the galactic bar are thought to
play an important role \citep[see][for a review of secular evolution in the
MW]{sellwood2014}.
Although the velocity of any individual star will only provide a weak age
constraint, the velocity dispersion of a group of stars can indicate whether,
on average, that group is old or young relative to other groups.
In this work we compare the velocity dispersions of groups of stars to
ascertain which groups are older and which younger and draw conclusions based
on the implied relative ages of populations.
% Stellar velocities have a history of being used as an age proxy, with several
% notable examples within stellar astronomy \citep[\eg][]{faherty2009,
% west2011}.
Since the age-velocity dispersion relations (AVRs) are themselves still
actively being calibrated, it is difficult to directly compare gyrochronal
ages with kinematic ones.
However, regardless of the exact relation between velocity dispersion and
stellar age, it is expected to be a monotonic relationship, therefore velocity
dispersion can be used effectively to {\it rank} groups of stars by age.

% Vertical {\it actions} are better age indicators than velocities, because
% actions are calculated by integrating angular momentum over the Milky Way's
% potential, and are therefore position invariant -- \eg\ a star will have the
% same action at periapsis and apoapsis.
% In contrast, orbital velocities are different at periapsis and apoapsis -- so
% in this sense, {\it actions} are the natural quantities to use as age proxies.
Vertical action is a better age indicator than vertical velocity
\citep[although still only a weak age indicator for individual
stars,][]{beane2018}, however both vertical action ($J_z$) and vertical velocity
(\vz/W) can only be calculated with full 6-dimensional position and velocity
information.
Unfortunately, most stars with measured rotation periods do not have radial
velocity (RV) measurements because they are relatively faint \kepler\ targets
($\sim$11th-18th magnitudes).
For this reason, we used velocity in the direction of galactic latitude, \vb.
The \kepler\ field is positioned at low galactic latitude (b=5-20\degrees), so
\vb\ is a close (although not perfect -- see section \ref{sec:results})
approximation to \vz.

@@

% Although only calibrated using Praesepe ($\sim$ 650 Myr) and the Sun (4.56
% Gyr), the \citet{angus2019} gyrochronology relation predicts accurate ages for
% members of NGC 6819, a 2.5 Gyr open cluster.
% However there are no rotation periods for K or M dwarfs in this cluster -- its
% coolest members with rotation periods are G dwarfs.
% The period-color relation of Praesepe is nearly identical to the period-color
% relation of the Hyades, a cluster of around the same age: $\sim$ 650 years
% \citep{douglas2016, douglas2017, rebull2017}.
% However, Praesepe and the Hyades do {\it not} have the same period-color
% relation as NGC 6811, a 1.1 Gyr cluster \citep{curtis2019}.
% The G dwarfs in NGC 6811 rotate at the same rate as the K dwarfs and it
% appears as though the K stars have `stalled' -- their spin-down has been
% halted.

% The \citet{angus2019} gyrochronology relation assumes that the rotation
% period-color relation of Praesepe is applicable to stars of all ages: the same
% polynomial relation fit to Praesepe is used to describe the period-color
% relation for all stars.
% However, the NGC 6811 cluster suggests that this assumption does not hold for
% K dwarfs past around 1 Gyr.
% If the period-color relation of the \citet{angus2019} gyrochronology model
% {\it were} a perfect model for the rotational evolution of stars, then groups
% of stars selected to be similar ages using this relation should fall on an
% isochrone on a CMD.
% % and have the same velocity dispersion across all colors.
% Unfortunately, uncertainties on \gaia\ photometry and parallaxes, plus
% variations in metallicity and extinction, blur out the main sequence enough
% that differences between stars in different age and rotation bins are not
% easily discernable on the CMD, although there is still a general age and
% rotation period gradient on the CMD, as seen in figures \ref{fig:CMD_cuts} and
% \ref{fig:age_gradient}.
% For this reason, we chose to use {\it kinematic} `isochrones', instead of
% magnitude and color-based isochrones: although kinematics as an age indicator
% is not necessarily as well calibrated to an absolute age scale as CMD
% position, it can be very sensitive to {\it differences} in the ages of stellar
% populations.

% Isochrones and stellar evolution tracks are highly dependent on choices made
% about input physics and assumptions.
% % and have often been calibrated using
% % different types of stars.
% As a result, different sets of models can have very different shapes on the
% CMD, particularly at low masses.
% In fact, only empirical models, not physical ones, are currently able to
% reproduce the CMD positions of M dwarfs.
% Instead of relying on CMD position to age-date groups of stars, we opted to
% explore age trends via kinematics.
% Kinematic age-dating has the advantage of being relatively model independent,
% or at least, having a very simple model: that velocity dispersion increases
% over time.
% This means that it is relatively easy to rank groups of stars by age: older
% groups have a larger velocity dispersion.

### 2019-08-19 paper/method.tex
redrafting

% 270 words
% Limit: 500 words
\section{Method}

\subsection{The data}

We crossmatched the \mct\ catalog of stellar rotation periods, measured from
\kepler\ light curves, with the \gaia\ DR2 catalog.
Reddening and extinction from dust was calculated for each star using the
Bayestar dust map implemented in the dustmaps {\it Python} package
\citep{green2018}.
We estimated effective temperatures from dereddened \Gaia\ \gcolor\ color,
using an 8th-order polynomial relation calibrated using .... stars
\racomment{ask Jason for details}.
\begin{equation}
    \mathrm{T_{eff}} = 8960 -4802C + 1931C^2 -2446C^3 + 2669C^4 - 1324C^5 +
    301C^6 - 26C^7,
% 8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
%           -2445.9980716705322, 2669.0248055458069, -1324.0671020746231,
%           301.13205924630165, -25.923997443169355]
\end{equation}
where C is \gaia\ \gcolor.

\begin{figure}
  \caption{
A \gaia\ color magnitude diagram showing the \citet{mcquillan2014} sample with
    extinction-corrected magnitudes, colored by rotation period.
We excluded photometric binaries and subgiants from our analysis by removing
stars above the two dashed lines.
% The rotation periods of binaries and subgiants do not follow a Skumanich-like
% braking law.
% The rotation period gradient across the main sequence is visible by eye in
%     this figure: young, rapidly rotating stars are located below the old,
%     slowly rotating stars.
% since we rely on a simple scaling
% between rotation period and age to make the argument that the rotation period
% is not caused by incorrect period measurements.
}
  \centering
    \includegraphics[width=1\textwidth]{CMD_cuts}
\label{fig:CMD_cuts}
\end{figure}
To explore the age of this stellar population from a rotation-period
standpoint, it was first necessary to remove visual binaries and subgiants
from the sample.
The rotation-period evolution of these two types of stars is generally
different to that of single stars which more usually follow a Skumanich-like
spin-down law.
% Tidal and magnetic interactions between the two components of a binary system
% can influence the rotation periods of both stars, and the expanding envelopes
% of subgiants drive rapid spin-down through conservation of angular momentum.
We removed visual binaries and subgiants from the sample by applying cuts to
the color-magnitude diagram (CMD), shown in figure \ref{fig:CMD_cuts}.
We fit a 6th-order polynomial to the main sequence and raised it by 0.22 dex,
to approximate the division between single stars and visual binaries.
We removed all stars above this line from the sample.
We also removed subgiants by eliminating stars brighter than 6th magnitude in
\gaia\ G-band.
\begin{figure}
  \caption{
Dereddened main sequence stars with \mct\ rotation periods on the \gaia\ CMD
    with visual binaries removed.  Points are colored by their gyrochronal
    age, according to the
    \citet{angus2019} gyrochronology relation.
    A general age gradient is visible across the main sequence.
}
  \centering
    \includegraphics[width=1\textwidth]{age_gradient}
\label{fig:age_gradient}
\end{figure}

We removed stars with negative parallaxes and parallax signal-to-noise ratios
below 10 from the sample.
We also removed a small number of stars fainter than 16th magnitude.
We used the {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy}
\citep{astropy2013, astropy2018} {\it Python} packages to calculate stellar
velocities.
{\tt Pyia} has built-in functionality for calculating velocity samples from
the full \gaia\ uncertainty covariance matrix via Monte Carlo sampling.
It therefore not only incorporates uncertainties on the \gaia\ positions
parallaxes and proper motions, it also accounts for the {\it covariance}
between these properties.
We removed stars with absolute \vb\ uncertainties greater than 1 \kms\ from
the sample.

>>>>

% 270 words
% Limit: 500 words

\subsection{The data}
\label{sec:the_data}

We used the publicly available \kepler-\gaia\ DR2 crossmatched
catalog\footnote{Available at gaia-kepler.fun} to combine the \mct\ catalog of
stellar rotation periods, measured from \kepler\ light curves, with the \gaia\
DR2 catalog.
Reddening and extinction from dust was calculated for each star using the
Bayestar dust map implemented in the {\tt dustmaps} {\it Python} package
\citep{green2018}, and {\tt astropy} \citep{astropy2013}.
We estimated effective temperatures from dereddened \Gaia\ \gcolor\ color,
using an 8th-order polynomial relation calibrated using .... stars
\racomment{ask Jason for details}.
\begin{equation}
    \mathrm{T_{eff}} = 8960 -4802C + 1931C^2 -2446C^3 + 2669C^4 - 1324C^5 +
    301C^6 - 26C^7,
% 8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
%           -2445.9980716705322, 2669.0248055458069, -1324.0671020746231,
%           301.13205924630165, -25.923997443169355]
\label{eq:curtis}
\end{equation}
where C is \gaia\ \gcolor.

% \begin{figure}
%   \caption{
% A \gaia\ color magnitude diagram showing the \citet{mcquillan2014} sample with
%     extinction-corrected magnitudes, colored by rotation period.
% We excluded photometric binaries and subgiants from our analysis by removing
% stars above the two dashed lines.
% % The rotation periods of binaries and subgiants do not follow a Skumanich-like
% % braking law.
% % The rotation period gradient across the main sequence is visible by eye in
% %     this figure: young, rapidly rotating stars are located below the old,
% %     slowly rotating stars.
% % since we rely on a simple scaling
% % between rotation period and age to make the argument that the rotation period
% % is not caused by incorrect period measurements.
% }
%   \centering
%     \includegraphics[width=1\textwidth]{CMD_cuts}
% \label{fig:CMD_cuts}
% \end{figure}
% To explore the age of this stellar population from a rotation standpoint, it
% was first necessary to remove visual binaries and subgiants from the sample.
% The rotational evolution of these two types of stars is generally different to
% that of single stars which more usually follow a Skumanich-like spin-down law.
% Tidal and magnetic interactions between the two components of a binary system
% can influence the rotation periods of both stars, and the expanding envelopes
% of subgiants drive rapid spin-down through conservation of angular momentum.
We removed visual binaries and subgiants from the sample by applying cuts to
the color-magnitude diagram (CMD), shown in figure \ref{fig:age_gradient}.
We fit a 6th-order polynomial to the main sequence and raised it by 0.27 dex
to approximate the division between single stars and visual binaries and
removed all stars above this line from the sample.
We also removed subgiants by eliminating stars brighter than 6th magnitude in
\gaia\ G-band.
\begin{figure}
  \caption{
Dereddened MS \kepler\ stars with \mct\ rotation periods on the \gaia\ CMD.
We excluded visual binaries by removing stars above the dashed line.
    Points are colored by their gyrochronal
    age, according to the
    \citet{angus2019} gyrochronology relation.
    A general age gradient is visible across the main sequence.
}
  \centering
    \includegraphics[width=1\textwidth]{age_gradient_combo}
\label{fig:age_gradient}
\end{figure}

The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
astropy2018} {\it Python} packages were used to calculate stellar velocities.
{\tt Pyia} has built-in functionality for calculating velocity samples from
the full \gaia\ uncertainty covariance matrix via Monte Carlo sampling.
It therefore not only incorporates uncertainties on the \gaia\ positions
parallaxes and proper motions, it also accounts for the {\it covariance}
between these properties.
We removed stars with negative parallaxes, parallax signal-to-noise ratios
less than 10, stars fainter than 16th magnitude, stars with absolute \vb\
uncertainties greater than 1 \kms\, and stars with galactic latitudes greater
than 15\degrees\ from the sample.
Because \vb\ is only a close approximation to \vz\ at low galactic latitudes,
and because latitude is correlated with stellar mass (lower mass stars are
older and tend to be preferentially located at high $b$), we found that
including stars with $b > 15$\degrees\ led to a larger number of \vb\ outliers
at low stellar mass.
15\degrees\ was found to be an adequate compromise between maintaining a close
relationship between \vb\ and \vz\ and keeping a large enough sample size to
provide meaningful results.
We also found that the anti-correlation between stellar mass and $b$ is
negligable for stars cooler than 5000 K.
For this reason, and because the rotational evolution of old, G stars does not
follow a simple Skumanich-like spin-down law \citep{vansaders2016}, we
restricted our sample to temperatures between \tmin\ and 5000 K, \ie\ K
dwarfs.

To explore the relationship between rotation period, \teff\ and age, we
selected groups of stars within different age ranges \citep[where age was
calculated using the][gyrochronology relation]{angus2019}, and calculated the
velocity dispersion: the standard deviation of velocities, \sigmavb, as a
function of effective temperature for each age group.
We performed 3$\sigma$ sigma-clipping on the stellar velocities in each age
and temperature bin to remove non-Gaussian outliers.
Without sigma-clipping, we found that a small number of high velocity
outliers at the low-temperature end of our sample substantially raised the
velocity dispersion for cooler stars.
Ages were calculated using dereddened \gaia\ \gcolor\ color, however
throughout this paper we show rotation periods as a function of {\it effective
temperature}.
We chose to use effective temperatures and not colors in this analysis as it
is the linear quantity and therefore easier to divide into bins of roughly
equal numbers of stars.
